### 2018-05-18 Writing/disparity_review.tex
Edits down to Future

\subsection{Tests}
Similarly to all the points covered above, the statistical test to apply solely depends on which question is tackled with disparity analysis (i.e. which hypothesis to test).
Considering the question, the data, the ordination method used (or the absence thereof) and the metric used, the statistical test should be based on a clearly state hypothesis.
This might seem obvious but we believe doing so greatly enhances the framing of the analysis and its comprehension by future readers.

For example, one commonly used test in morphological analysis, is the non parametric permutation anovas [NPANOVA or PERMANOVA; CITE Anderson + examples].
This test and its implementations is perfectly valid to test whether two groups have a different variance/covariance but is based on the pairwise distances of the group elements.
This test will ignore any eventual ordination (if the data was ordinated) and is not based on a uni-dimensional disparity metric.
In other words, if one observes no overlap between two groups on the two first axis of a PCA, the hypothesis H0 that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance based in a ``distance-space''.
Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the disparity analysis question.

Another caveat for applying statistical tests on disparity data is on which data to actually apply the test.
In fact, in morphological disparity analysis (especially for palaebiological questions), data is often bootstrapped.
This has two useful advantages: first, when the disparity metric is unidimensional (e.g. the sum of variances or the average squared pairwise distance), bootstrapping the data will generate a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit to compare distributions; second, when the data is scarce (e.g. for palaeontological data), bootstrapping the data will allow to introduce statistical variance and will render the test less sensitive to outliers.
However, bootstrapping the data also comes with an important caveat: the data analysed is pseudo-replicate and thus non-independent.
This can violate most of the parametric statistical tests assumptions.
Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the type I error.

>>>>

\subsection{Tests}
The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.

For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.

Another caveat for applying statistical tests on disparity data is that it is important to consider which data to apply the test on.
For example, in morphological disparity analysis (especially for palaeobiological questions), data is often bootstrapped.
This has two useful advantages: first, when the disparity metric is unidimensional (e.g. the sum of variances or the average squared pairwise distance), bootstrapping the data generates a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit available for comparing distributions; second, when the data is scarce (e.g. for palaeontological data), bootstrapping the data allows users to introduce statistical variance, rendering the test less sensitive to outliers.
However, bootstrapping the data also comes with an important caveat: the data analysed are pseudo-replicated and thus non-independent.
This can violate the assumptions of parametric statistical tests.
Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the Type I error.
